Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-07-16 Chapter_1/04_Manuscript/Manuscript.tex
Rewriting introduction

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
% Dispersal & Connectivity
Dispersal is an important process that allows species to avoid inbreeding,
colonize new habitats and reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and preservation of dispersal corridors has become a
task of extraordinary importance for conservation authorities.

% Quantifying Connectivity
Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their usefulness in reality. Individual-based dispersal simulations have
been proposed to address these shortcomings, yet due to the sheer amount of
non-trivial modeling decisions decisions, a unified and objective framework to
simulate dispersal is missing.

% Step Selection Functions
Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful to
investigate \textit{transient} movement behavior during dispersal and landscape
connectivity.

% Current Study
Here, we showcase the use of integrated step selection functions as simple,
individual-based, and spatially explicit movement model to simulate dispersal of
the endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions, thereby parametrizing a fully
mechanistic movement model describing how dispersing wild dogs move through the
available landscape. Based on this model, we simulate 80'000 dispersers moving
across the extent of the KAZA-TFCA and generate a set of three connectivity
maps, each focused on a different aspect of connectivity. Finally, we discuss
the benefits and pitfalls of such a simulation-based approach and highlight
potential improvements to be made in the future.

@@

% Create linenumbers
% \linenumbers

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the social structure of wild animal populations that are
distributed in space \citep{Hanski.1998, Clobert.2012} and may strongly affect
population dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to maintain genetic diversity
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013}, to
rescue small, non-viable populations \citep{Brown.1977}, and to promote the
colonization or recolonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
landscape connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become a
fundamental task in conservation science \citep{Nathan.2008, Doerr.2011,
Rudnick.2012}. The ability to pinpoint relevant dispersal hotspots requires
information on movement behavior during dispersal and knowledge about factors
that limit dispersal and therefore connectivity \citep{Baguette.2013,
Vasudev.2015}.

% Advancements in GPS Technology & Movement Ecology
\subsection{Advancements in GPS Technology \& Movement Ecology (90\%)}
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the use of GPS data to study animal dispersal and
connectivity has accelerated \citep{Elliot.2014, Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded a ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the availability of large amounts of
empirical data and an increased computational power have led to the development
of numerous techniques to study dispersal and highlight critical corridors
between subpopulations \citep{Boyce.2002, Fortin.2005, Cushman.2010,
Zeller.2012, Diniz.2020}.

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory (CT;
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape to infer likely movement corridors. Despite their
intuitive nature and ease of use, both methods make rigorous assumptions about
animal movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be fulfilled by
migrating animals that typically move between a discrete set of habitats through
familiar landscapes. Dispersers, on the other hand, usually move over long
distances into unknown territory and are therefore less likely to be aware of
associated movement costs \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Another
issue of LCPs analysis concerns the fact that least-costly routes, by their very
nature, are only one pixel wide \citep{Pinto.2009}. This neglects the fact that
alternative routes with similar costs may exist and implies that the width of
inferred movement routes depends on the resolution of chosen covariate layers
and may not be biologically meaningful \citep{Diniz.2020}. Although some of
these deficiencies can be addressed using less stringent versions of the least
cost algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness remains.

% Issues with Circuit Theory
\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are rarely ever met. For
example, because CT only allows movements from a source cell to its four or
eight adjacent cells, it implicitly posits that animals exhibit a perceptual
range of a single pixel. Given that covariate layers are usually resolved with a
pixel size between 30 m x 30 m and 1 km x 1 km, this hardly ever renders the
true capability of animals to perceive the environment. Moreover, CT is built
around the assumption of a complete random walk \citep{Diniz.2020}, entailing
that directional biases cannot be rendered. Nevertheless, directionality is a
common characteristic in animal movement \citep{Bovet.1991, Schultz.2001},
especially in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}.

% Issues with Both Methods
\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal \citep{Diniz.2020}. Statements about the expected
duration required to traverse a certain corridor are therefore impossible.
Likewise, because movement is not modeled explicitly, interactions between
movement and habitat preferences of the focal species cannot be rendered.
Connectivity therefore merely arises in result to the landscape structure, which
is usually referred to as structural connectivity. While structural connectivity
yields insights in the \textit{potential} of the landscape to be traversed, it
does not enable to quantify the \textit{actual} gene flow through the area.
Consequently, a functional view on connectivity, which also renders the
behavioral response of the animal with respect to prevailing habitat conditions,
is often more desirable \citep{Tischendorf.2000, Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal trajectories are simulated spatially explicitly, based on movement
rules that determine how individuals move over and interact with the prevailing
landscape \citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004,
Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020, Vasudev.2021}. Using the simulated trajectories,
one can calculate a set of connectivity metrics, such as inter-patch
connectivity and traversal frequency, to reveal major dispersal corridors
\citep{Kanagaraj.2013, BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}.
However, while IBMMS can be employed to overcome any of the shortcomings
intrinsic to LCPs and CT, they are subject to a vast amount of subjective,
non-trivial modeling decisions. Moreover, they can be challenging to fit and
require vast amounts of movement data, ideally collected during dispersal
\citep{Diniz.2020}. Consequently, alternative methods that require fewer
modeling decisions and are straight forward to apply are desirable.

% Step Selection Analysis
\subsection{Step Selection Analysis (90\%)}
Here, we investigate the usefulness of integrated step selection functions
(ISSFs, \citealp{Avgar.2016}), as a relatively simple but powerful IBMM based on
which dispersal can be simulated. While regular SSFs were intended to learn
about relative habitat preferences of the focal species \citep{Fortin.2005,
Thurfjell.2014, Avgar.2017}, the method has recently been generalized to
\textit{integrated} SSFs and now enables to jointly study habitat and movement
preferences, as well as potential interactions between them \citep{Avgar.2016,
Signer.2017, Fieberg.2020}. ISSFs therefore provide a relatively simple means to
model complex movement behavior, where movement is viewed as the result of two
intertwined behavioral kernels (e.g. \citealp{Prokopenko.2017, Munden.2020}).
Importantly, a parametrized ISSF model can be employed as a fully mechanistic
movement model based on which individual movement trajectories can be simulated
\citep{Avgar.2016, Signer.2017}. In fact, \cite{Signer.2017} used ISSF to
simulate steady state utilization distributions of resident animals that were
moving around a point of attraction. However, the degree to which such
simulations are helpful in detecting movement corridors and modeling landscape
connectivity remains to be investigated.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012}. As of
today, only 6'000 free-ranging individuals remain in small and spatially
scattered subpopulations \citep{Woodroffe.2012}. Within those subpopulations,
wild dogs form cohesive packs comprising 8 to 12 adults and their offspring
\cite{McNutt.1995}. After reaching sexual maturity, male and female offspring
form same-sex coalitions and disperse from their natal pack in search for
potential mating partners and a suitable territory to settle \citep{McNutt.1996,
Behr.2020}. New packs are formed when dispersing coalitions join unrelated
opposite-sex dispersing coalitions \citep{McNutt.1996}. Dispersing wild dogs can
cover several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
few strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area has originally been intended to facilitate migration of elephants, but is
expected to provide benefits to a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a basic habitat selection model and predicted a
permeability surface that we used to compute least-cost paths and corridors. We
now expand on this knowledge and use ISSFs to develop a more mechanistic
movement model of dispersing wild dogs (\Cref{GraphicalAbstract}). We employ the
model to simulate dispersers moving across the KAZA-TFCA and generate three
distinct connectivity maps, each shedding light onto a different aspect of
connectivity. With this work, we exemplify how ISSFs can be utilized for
dispersal simulations and we discuss several benefits of this approach over
traditional connectivity modeling techniques such as least-cost analysis and
circuit theory. Most importantly, simulations based on ISSFs provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addition, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis.

@@

\subsection{GPS Relocation Data (90\%)}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana \citep{Cozzi.2020, Hofmann.2021}. We selected potential
dispersers based on age, pack size, number of same‐sex siblings within the pack,
and presence of unrelated opposite-sex individuals in the pack
\citep{McNutt.1996, Behr.2020}. We immobilized selected individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020} that was
injected by dart, fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Immobilized individuals were fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}) that guaranteed automated drop-off through a decomposable piece of
cotton. Handling and collaring of all individuals was supervised by a
Botswana-registered wildlife veterinarian and all individuals quickly rejoined
their pack after immobilization.

16 collared individuals eventually dispersed, each in a separate same-sex
dispersal coalition (7 female and 9 male coalitions). During dispersal, collars
were programmed to record a GPS fix every 4 hours, all of which were regularly
transmitted over the Iridium satellite system, thereby allowing to remotely
track individuals, even if they left the main study area and crossed
international borders. Because behavior during dispersal is more pertinent for
assessing landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded
all data that was collected during residency and only retained GPS data recorded
during dispersal. In some instances, exact dispersal dates were known from field
observations, whereas in other cases we determined dispersal phases using the
net-squared displacement metric. Net squared displacement measures the squared
Euclidean distance of a GPS relocation to a reference point \citep{Borger.2012},
which in our case was set to the center of each individual's natal home range.
Thus, dispersal was deemed to have started when an individual left its natal
home range and ended once individuals became sedentary again. As previous
research found no differences in behaviors of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by an individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not successfully collected
on the 4-hourly schedule (\( \pm \) 15 minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting water layers thus temporally aligned with our
dispersal events. We furthermore computed a proxy for human influence, rendering
anthropogenic pressures stemming from human-density, agricultural sites, and
roads. All spatial layers were coarsened or interpolated to a target resolution
of 250 m by 250 m. Further details on the sources and preparation of each
habitat covariate are given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern \citep{Castello.2018}, we also coded
a binary variable (\textsf{LowActivity}) indicating whether a step was realized
during periods of low wild dog activity (17:00 to 09:00 local time) or high wild
dog activity (09:00 to 17:00 local time). Handling and manipulation of all data,
as well as all models and simulations were implemented with the statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed step plus its 24 random steps formed a stratum
and received a unique identifier. As suggested by \cite{Avgar.2016}, we
generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale = 6'308, shape = 0.37). Along each step, we
extracted and averaged spatial covariates using the {\tt velox} package
\citep{Hunziker.2021}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each observed and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| > 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

@@

Assuming that all other covariates are held constant at their means, the habitat
kernel reveals that dispersing wild dogs avoid water but prefer its proximity.
Similarly, dispersers avoid areas that are covered by woodlands, yet prefer
regions covered by shrublands or grasslands. Finally, dispersers avoid movement
through landscapes that are dominated by humans. Effect sizes are strong and,
except for effect of \textsf{distance to water}, statistically clear on the 5\%
significance level.

>>>>

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection
function, Kavango-Zambezi Transfrontier Conservation Area, landscape
connectivity, Lycaon pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

%------------------------------------------------------------------------------
% Short Version
%------------------------------------------------------------------------------
% The ability to disperse is contingent a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists mainly rely on least-cost analysis or circuit
% theory to investigate connectivity, albeit both methods make several assumptions
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we employ a simple three-step workflow that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In the first step, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In a
% second step, we apply the parametrized model as an individual-based movement
% model to simulate dispersal trajectories. Finally, in the third step, we
% consolidate simulated trajectories into three complementary connectivity maps,
% each focusing on a different aspect of landscape connectivity.
%
% We apply these steps to assess landscape connectivity for dispersing African
% wild dogs (\textit{Lycaon pictus}) southern Africa. We thereby demonstrate that
% simulations from integrated step selection functions offer a simple, yet
% powerful alternative to traditional connectivity modeling techniques.

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of Dispersal & Connectivity
Dispersal of individuals is a crucial pre-requisite for long-term species
persistence. However, the ability to disperse is contingent a sufficient degree
of landscape connectivity, which is why the identification and preservation of
movement corridors that promote connectivity has become a task of exceptional
importance.

% Quantifying connectivity
Over the past two decades, ecologists have primarily relied on least-cost
analysis and circuit theory to investigate connectivity. Unfortunately, both
methods make several restrictive assumptions that limit their usefulness in
reality. Least-cost analysis, for instance, assumes that animals move towards a
pre-conceived endpoint and choose a cost-minimizing route accordingly. Circuit
theory, in contrast, presumes a random walk without any directional persistence.
To address these shortcomings, simulations from individual-based movement models
have been proposed and applied, yet a unified framework to simulate dispersal
and to derive connectivity maps based on simulated dispersal events is lacking.

% What we do
% and synthesize a simple
% three-step workflow that uses
Here, we combine several already-existing methods to simulate dispersal and
investigate landscape connectivity in three distinct steps. In the first step,
integrated step selection functions are used to parametrize a mechanistic
movement model. In the second step, the parametrized model is applied as an
individual-based movement model to simulate dispersal trajectories. In the third
step, simulated trajectories are consolidated and synthesized into three
complementary connectivity maps, each focusing on a different aspect of
landscape connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors,
and a map of inter-patch connectivity, depicting the presence or absence of
functional links between distinct habitat patches.

% Case study
As a case study, we apply the proposed workflow to assess landscape connectivity
for dispersing African wild dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi
Transfrontier Conservation Area, the world's largest transboundary conservation
area. With this, we demonstrate that the outlined workflow offers a simple, yet
powerful alternative to traditional connectivity modeling techniques. Although
such individual-based simulations necessitate informed decisions about the
number of simulated individuals and the duration of simulated dispersal events,
they enable to relax several of the unrealistic assumptions inherent to
least-cost analysis or circuit theory.

@@

% Create linenumbers
\linenumbers

@@

% Importance of Dispersal & Connectivity
\subsection{Dispersal \& Connectivity Models (90\%)}
Successful dispersal of individuals requires a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become a fundamental task in
conservation science \citep{Nathan.2008, Doerr.2011, Rudnick.2012}. To pinpoint
relevant dispersal corridors, a comprehensive understanding of the factors that
limit dispersal of the focal species and an adequate connectivity model are
required \citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two
most prominent connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based approaches that estimate conductance of the
landscape based on predicted habitat permeability \citep{Zeller.2012}.

% Issues with Both Methods
\subsection{Issues with Traditional Methods (90\%)}
Despite their intuitive nature and ease of use, both methods make several
assumptions that are hardly ever met in reality \citep{Diniz.2020}. With LCPA,
for instance, a least costly route always exists, even if associated movement
costs are unreasonably high and will never be incurred by a dispersing
individual. The method also presumes that animals move towards a pre-conceived
endpoint and choose a cost-minimizing route accordingly. While these assumptions
may by fulfilled by migrating animals, dispersers usually move into unfamiliar
territory and are therefore less likely to be aware of associated movement costs
\citep{Koen.2014, Abrahms.2017, Cozzi.2020}. CT relaxes these assumptions, yet
posits that animals move according to a random walk, entailing that directional
biases cannot be rendered \citep{Diniz.2020}. In reality, however, directional
persistence is regularly observed \citep{Bovet.1991, Schultz.2001}, especially
in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}. Some of these
deficiencies can be addressed using alternative graph-based approaches, such as
least-cost \textit{corridors} \citep{Pinto.2009}, \textit{thresholded}
least-cost paths \citep{Landguth.2012}, or \textit{randomized} least-cost paths
\citep{Panzacchi.2016}, yet a certain degree of arbitrariness remains. Finally,
none of the methods is capable of rendering the temporal dimension of dispersal
\citep{Diniz.2020}, meaning that statements about the expected duration required
to traverse a certain corridor are impossible.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To overcome the unrealistic assumptions inherent to traditional connectivity
modeling techniques, simulations from individual-based movement models (IBMMs)
have been proposed and applied \citep{Diniz.2020}. In these models, dispersal
trajectories are simulated based on movement rules that determine how
individuals move across or interact with the prevailing landscape
\citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004, Revilla.2004,
Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019,
Zeller.2020, Vasudev.2021}. Instead of enforcing connections towards a known
endpoint, IBMMs allow individuals to choose their route as they move and enable
to render potential interactions between movement behavior and environmental
conditions. IBMMs furthermore enable to model directional persistence and to
study the temporal dimension of dispersal \citep{Diniz.2020}. Despite these
advantages, however, a unified framework to simulate dispersal and to derive
connectivity maps based on simulated dispersal events is lacking.

% The Power of Step Selection Functions
\subsection{Step Selection Functions (90\%)}
A promising framework that has been developed more recently...
Over the past decade, the increased availability of large amounts of data and an
increased computational power have led to the development of numerous techniques
to study animal movement and resource selection during dispersal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012}. Especially
\textit{resource selection functions} \citep{Boyce.2002}, \textit{step selection
functions} \citep{Fortin.2005}, and \textit{path selection functions}
\citep{Cushman.2010} have pushed the field of movement ecology forward. These
methods allow estimating habitat preferences of the focal species by comparing
covariates at locations visited by the animal to the same covariates at
locations available to, but not visited by the animal \citep{Boyce.2002,
Fortin.2005, Cushman.2010, Thurfjell.2014}. While regular SSFs were intended to
learn about relative habitat preferences of the focal species
\citep{Fortin.2005, Thurfjell.2014, Avgar.2017}, the method has recently been
generalized to \textit{integrated} SSFs and now enables to jointly study habitat
and movement preferences, as well as potential interactions between them
\citep{Avgar.2016, Signer.2017, Fieberg.2020}. ISSFs therefore provide a
relatively simple means to model complex movement behavior, where movement is
viewed as the result of two intertwined behavioral kernels (e.g.
\citealp{Prokopenko.2017, Munden.2020}). Importantly, a parametrized ISSF model
represents a fully mechanistic movement model based on which individual movement
trajectories can be simulated \citep{Avgar.2016, Signer.2017}. In fact,
\cite{Signer.2017} used ISSF to simulate steady state utilization distributions
of resident animals that were moving around a point of attraction. However, the
degree to which such simulations are helpful in detecting movement corridors and
modeling landscape connectivity remains to be investigated.

% What we do
\subsection{Proposed Framework (90\%)}
Here, we therefore propose to combine several already-existing methods into a
simple three-step workflow to simulate dispersal and investigate landscape
connectivity. In the first step, integrated step selection functions are used to
parametrize a mechanistic movement model rendering dispersal behavior. In the
second step, the parametrized model is applied as an individual-based movement
model to simulate dispersal trajectories. Finally, in the third step, simulated
trajectories are consolidated into three complementary connectivity maps, each
focusing on a different aspect of landscape connectivity. The set of maps
includes a heatmap, revealing frequently traversed areas, a betweenness-map,
delineating critical dispersal corridors, and a map of inter-patch connectivity,
depicting the presence or absence of functional links between distinct habitat
patches.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
We exemplify application of the proposed workflow by investigating landscape
connectivity for the endangered African wild dog (\textit{Lycon pictus}). While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range, mainly due to persecution by humans,
habitat fragmentation and destruction, and deadly diseases
\citep{Woodroffe.2012}. Within the few remaining subpopulations, wild dogs form
cohesive packs comprising 8 to 12 adults and their offspring \cite{McNutt.1995}.
After reaching sexual maturity, male and female offspring form same-sex
coalitions and disperse from their natal pack in search for potential mating
partners and a suitable territory to settle \citep{McNutt.1996, Behr.2020}. New
packs are formed when dispersing coalitions join unrelated opposite-sex
dispersing coalitions \citep{McNutt.1996}. During dispersal, wild dogs can cover
several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
main strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area was originally devoted towards facilitating the migration of elephants, yet
it can be expected to yield benefits to a great multitude of other species
\citep{Elliot.2014, Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
We previously assessed landscape connectivity for dispersing African wild dogs
within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}. More
specifically, we fitted a habitat selection model and predicted a permeability
surface based on which we computed least-cost paths and corridors between
protected areas. We now expand on this knowledge and use ISSFs to develop a more
detailed, mechanistic movement model of dispersing wild dogs
(\Cref{GraphicalAbstract}). We then employ the model to simulate 80'000
dispersers moving 2'000 steps across the KAZA-TFCA. Finally, we generate a
heatmap, betweenness map and a map of interpatch connectivity. We thereby
exemplify how ISSFs can be utilized for dispersal simulations and we discuss
several benefits of this approach over traditional connectivity modeling
techniques such as least-cost analysis and circuit theory. Most importantly,
simulations based on ISSFs provide a more generic view on how connectivity
emerges and to which degree connectivity depends on the dispersal duration. In
addition, by generating proper dispersal trajectories, network theory can be
applied to calculate network metrics that are pertinent to connectivity
analysis.

@@

\subsection{GPS Relocation Data (90\%)}
GPS data of 16 dispersing wild dogs was obtained as described in
\citep{Cozzi.2020, Hofmann.2021}. We collected data on 7 female and 9 male
dispersing coalitions between 2011 and 2019 from a free-ranging population
inhabiting the Moremi National Park in northern Botswana. In some instances,
exact dispersal dates were known from field observations, in other cases we
determined dispersal phases using the net-squared displacement metric. Net
squared displacement measures the squared Euclidean distance of a GPS relocation
to a reference point \citep{Borger.2012}, which in our case was set to the
center of each individual's natal home range. Thus, dispersal was deemed to have
started when an individual left its natal home range and ended once individuals
became sedentary again. Because behavior during dispersal is more pertinent to
landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data
that was collected during residency. During dispersal, collars were programmed
to record a GPS fix every four hours and regularly transmitted over the Iridium
satellite system. As previous research found no differences in behaviors of
females and males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did
not distinguish between sexes. After collection, we converted GPS coordinates (n
= 4'169) to steps, where each step represented the straight-line distance
traveled by an individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule (\( \pm \) 15
minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting layers thus temporally aligned with each
dispersal event, ensuring minimal temporal lag between observed movements and
corresponding water layer. We furthermore computed a proxy for human influence,
rendering anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All spatial layers were coarsened or interpolated to a target
resolution of 250 m by 250 m. A detailed description of the preparation of each
habitat covariate is given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern \citep{Castello.2018}, we also coded
a binary variable (\textsf{LowActivity}) indicating whether a step was realized
during periods of low wild dog activity (17:00 to 09:00 local time) or high wild
dog activity (09:00 to 17:00 local time). Handling and manipulation of all data,
as well as all models and simulations were implemented with the statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps, such that a realized step plus its 24 random steps formed
a stratum that received a unique identifier. As suggested by \cite{Avgar.2016},
we generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale = 6'308, shape = 0.37). Along each step, we
extracted and averaged spatial covariates using the {\tt velox} package
\citep{Hunziker.2021}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each observed and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| > 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

@@

Assuming that all other covariates are held at their means, the habitat kernel
reveals that dispersing wild dogs avoid water but prefer its proximity.
Similarly, dispersers avoid areas that are covered by woodlands, yet prefer
regions covered by shrublands or grasslands. Finally, dispersers avoid movement
through landscapes that are dominated by humans. Effect sizes are strong and,
except for effect of \textsf{distance to water}, statistically clear on the 5\%
significance level.
